\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization \thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602 (phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}

### 2009-12-29 honors_thesis/journal_paper/paper.tex
Started working on the journal version.  It is coming along

\documentclass[conference,letterpaper]{IEEEtran}

@@

The original PSO can be reproduced exactly, two iterations at a time, using
this technique.  However, in strictly using speculative evaluation to reproduce
the original PSO, a lot of positions are speculatively evaluated whose values
get ignored after deciding which branch was the correct one.  A better use of
those speculative evaluations may be to incorporate the information into the
swarm, no longer exactly reproducing PSO but perhaps producing better results.

\sec{pso} describes the particle swarm optimization algorithm.  \sec{sepso}
then describes how speculative evaluation can be done in parallel PSO to
perform two iterations at once.  In \sec{extra}, we discuss a variant of the
speculative algorithm which uses all of the speculative evaluations performed
instead of throwing away all but one for each particle.  We then show in
\sec{results} a comparison between the original PSO algorithm, our speculative
algorithm, and its variant.  In \sec{conclusion} and \sec{future} we conclude
and discuss future work.

@@

\section{Using All Speculative Evaluations}
\label{sec:extra}

In performing speculative evaluation as we have described it, $2(n+1)$
speculative evaluations are done per particle, while all but one of them are
completely ignored.  It seems reasonable to try to make use of those
evaluations instead of throwing them away.  Making use of them changes the
behavior of PSO, instead of reproducing it exactly as the above method
explains, but the change could be an improvement.

@@

@@

In our experiments we compared our speculative PSO algorithm to the standard
PSO algorithm.  To use the same number of processors for each algorithm,
speculative evaluation needs a smaller swarm size than standard PSO.  For each
of the topologies we used, a particle has two neighbors other than itself.  As
shown in Table~\ref{tab:evals}, this results in $7$ speculative evaluations per
particle.  With one evaluation needed for the original, non-speculative
particle, we have $8p_s$ evaluations for every two iterations, where $p_s$ is
the number of particles in the speculative swarm.  The extra evaluations done
in speculative evaluation would instead be particles in regular PSO, so we
compare swarms of size $p_s$ in speculative evaluation with swarms of size
$8p_s$ in regular PSO.

We use a Random topology in our speculative algorithm, as the Complete topology
leads to an explosion of speculative evaluations.  However, if speculative
evaluation were not being performed, it is likely that the Complete topology
would be used, ignoring the communication cost of Complete over Random.  To be
fair in our comparisons, for the standard PSO algorithm we compare to both the
Random topology with two neighbors and to the Complete topology.  All functions
have 20 dimensions, and all results shown are averages over 20 runs.  We use
240 processors in our experiments, so speculative algorithms use 30 particles
and standard PSO uses 240 particles.  For ease of labeling, we call our
speculative algorithm SEPSO, the variant SEPSO Pick Best, and standard PSO just
PSO.

First we look at the function Sphere, the simplest of common benchmark
functions.  Its equation is $f(\Vec{x}) = \sum_{i=1}^D x_i^2$.  The function
has no local optima and is best optimized in terms of function evaluations with
a small swarm using a Complete topology.  Accordingly, we show results for a
small Complete swarm in \fig{sphere1}.  In comparing SEPSO with PSO, we can see
that SEPSO clearly beats PSO with a Random topology.  However, the performance
between SEPSO and PSO Complete is not as encouraging---PSO Complete performs
slightly better.  But SEPSO Pick Best handily outperforms all other algorithms
in this problem.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere1.eps}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations.  Speculative algorithms have a swarm of 30 particles and do two
  iterations per round, while standard PSO algorithms use 240 processors and do
  one iteration per round.}
  \label{fig:sphere1}
\end{figure}

Next we look at the function Griewank, defined by the equation $f(\Vec{x}) =
\frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
\right) + 1$.  It is generally recommended to use the Ring topology when
optimizing the Griewank function, as Complete is prone to premature convergence
on a local optima.  We show results in \fig{griewank1} for swarms of size 30
and 240 using the Ring topology. One can see in the figure that both SEPSO
algorithms initially outperformed PSO, but got stuck in a local optima and were
eventually outperformed themselves.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank1.eps}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations.  Speculative algorithms have a swarm of 30 particles and do two
  iterations per round, while standard PSO algorithms use 240 processors and do
  one iteration per round.}
  \label{fig:griewank1}
\end{figure}

\fig{griewank1} is slightly deceptive, however, as with Griewank some of the
runs found the global optimum, 0, while some did not.  A more accurate plot for
this function is one showing what percentage of the runs have found the global
optimum at each iteration.  \fig{griewank2} shows such a plot.  A swarm of size
30 performing the PSO algorithm gets stuck a little less than half of the time,
while 240 particles provides enough exploration to find the global optimum
every time.  Surprisingly, using the SEPSO Pick Best algorithm with 30
particles greatly improves the chance of success with Griewank, while also
finding the global optimum in far fewer rounds of evaluations than standard
PSO.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank2.eps}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations.  Instead of showing average function value, we show the
  percentage of runs that have found the global optimum by each iteration.}
  \label{fig:griewank2}
\end{figure}

Lastly we look at a function that does not show much or any improvement with
this method.  The function Rastrigin is defined as $f(\Vec{x}) = \sum_{i=1}^D
\left(x_i^2 - 10\cos\left(2\pi x_i\right) + 10\right)$.  It has been shown that
with Rastrigin, the more particles there are in the swarm, the lower function
value it finds, up to at least 4000 particles~\cite{mcnabb-cec09}.  Smaller
swarms get caught in local optima.  Because our speculative algorithms have a
significantly smaller swarm size, they get stuck at higher values while the
larger swarms performing regular PSO continue to improve the best value found.
\fig{rastrigin} shows the results graphically.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{rastrigin.eps}
  \caption{Function Rastrigin with a swarm that uses 240 processors per round
  of evaluations.  Speculative algorithms have a swarm of 30 particles and do
  two iterations per round, while standard PSO algorithms use 240 processors
  and do one iteration per round.}
  \label{fig:rastrigin}
\end{figure}

@@

We have described how parallel implementations of particle swarm optimization
can be modified to allow additional processors to increase the number of
iterations of the algorithm performed, instead of merely adding more particles
to the swarm.  We have detailed several different implementations of the
algorithm, depending on the parallel PSO framework being used.  Using our
modifications, the original PSO algorithm is exactly reproduced two iterations
at a time.  This technique requires more function evaluations per iteration
than regular PSO, but for some functions still performs better when run in a
parallel environment.

We have also described a method for making use of extra speculative evaluations
that performs incredibly well on some functions.

Though space has limited our presentation of results to a very small sample of
the interesting and relevant applications of our technique, we believe the
results shown are sufficient motivation for the further exploration of
speculative evaluation in PSO.

Our results show that performing speculative evaluation instead of increasing
swarm size is only beneficial for some functions at the swarm sizes we used.
However, we would venture that for all functions there is some point at which
increasing the swarm size provides relatively little benefit.  At that point,
additional processors should be used to speed up the algorithm instead of add
additional particles.

There are some functions for which almost no exploration needs to be done;
Sphere is an example of such a function, as would be a function where the
general area of the optimal point is known, but the exact point needs to be
found.  For such functions the best use of processors is to have a small swarm
performing speculative evaluation with our second method, where all speculative
evaluations are used.

There are other functions for which there is a point where the algorithm has
done ``enough'' exploration and consistently finds the global optimum.  Adding
additional particles to the swarm is akin to increasing its capacity for
exploration.  We have seen that for a 20-dimensional Griewank function, a
number of particles greater than 30 but less than 240 has ``enough''
exploration and finds the optimal value every time.  For these functions, one
should use processors to increase the swarm size until enough exploration has
been reached, then use additional processors to increase the number of
iterations performed.

There are still other functions for which it seems there is never enough
exploration, such as the Rastrigin function.  It has been shown that up to 4000
particles there is no point at which ``enough'' exploration has been
done~\cite{mcnabb-cec09}.  For those functions the correct use of processors
cannot be easily prescribed and depends largely on the desired outcome.  It is
likely that the best thing to do with those functions is to use all processors
to add particles to the swarm and do no speculative evaluation.

@@

This work represents the first attempt of which we are aware at using
additional processors in parallel with PSO to do something other than either
add particles to the swarm or parallelize the evaluation of the objective
function.  We anticipate that many future advancements can be made.  For
example, our technique required speculatively evaluating every possible next
position of each particle.  The number of speculative evaluations could
potentially be greatly reduced by using statistics for which branch of motion
particles most often follow.  Only the subset of the most probable next
locations would then be evaluated.  Using an asynchronous PSO variant may also
be helpful in this case, allowing particles to be at different iterations in
case the correct branch was not guessed and needs to be evaluated separately.

Another possible advancement would be to reduce communication costs by dropping
some of the inter-processor messages in distributed frameworks.  The PSO
algorithm would have to be adapted to work with less information, perhaps not
requiring that each processor have completely accurate information about what
neighboring particles are doing.

Still another possibility is to view the set of possible evaluations as an
infinite tree.  Each level in the tree is an iteration, and the branches on
each level represent the possible next positions of the particle.  One need not
simply evaluate all of one level for each particle---perhaps going two or three
iterations ahead on the most likely path would perform better.

>>>>

\documentclass[journal,letterpaper]{IEEEtran}

@@

The original PSO can be reproduced exactly, two iterations at a time, using
this technique.  However, if we relax the requirements of the algorithm, no
longer demanding that it strictly mimic the behavior of the original PSO, we
can introduce new speculative techniques that often out-perform PSO.

\sec{pso} describes the particle swarm optimization algorithm.  \sec{sepso}
then describes how speculative evaluation can be done in parallel PSO to
perform two iterations at once, then presents a comparison of the speculative
algorithm with the original PSO.  In \sec{relax}, we discuss various methods of
improving the performance of speculative evaluation in PSO, all of which break
the requirement of strictly reproducing the behavior of the original algorithm.
In \sec{conclusion} and \sec{future} we conclude and discuss future work.

@@

\subsection{Comparison with Standard PSO}

We ran experiments to compare our speculative PSO algorithm to the standard
PSO algorithm.  To use the same number of processors for each algorithm,
speculative evaluation needs a smaller swarm size than standard PSO.  For each
of the topologies we used, a particle has two neighbors other than itself.  As
shown in Table~\ref{tab:evals}, this results in $7$ speculative evaluations per
particle.  With one evaluation needed for the original, non-speculative
particle, we have $8p_s$ evaluations for every two iterations, where $p_s$ is
the number of particles in the speculative swarm.  The extra evaluations done
in speculative evaluation would instead be particles in regular PSO, so we
compare swarms of size $p_s$ in speculative evaluation with swarms of size
$8p_s$ in regular PSO.

We use a Random topology in our speculative algorithm, as the Complete topology
leads to an explosion of speculative evaluations.  If speculative evaluation
were not being performed, it is possible that the Complete topology would be
used.  However, the Complete topology also requires a very large amount of
interprocessor communication in parallel PSO, so it is still quite possible
that Random would be used even with standard PSO.  But, to be fair in our
comparisons, we compare to the standard PSO algorithm using both the Random
topology with two neighbors and the Complete topology.  All functions have 20
dimensions, and all results shown are averages over 20 runs.  We use 240
processors in our experiments, so our speculative algorithm uses 30 particles
and standard PSO uses 240 particles.  For ease of labeling, we call our
speculative algorithm SEPSO and standard PSO just PSO, with the topology
following.

First we look at the function Sphere, the simplest of common benchmark
functions.  Its equation is $f(\Vec{x}) = \sum_{i=1}^D x_i^2$.  The function
has no local optima and is best optimized in terms of function evaluations with
a small swarm using a Complete topology.  Accordingly, we show results for a
Complete swarm in \fig{basic-sphere}.  In comparing SEPSO with PSO, we can see
that when both are using a Random topology, SEPSO clearly outperforms PSO.
However, the comparison between SEPSO and PSO Complete is not as
encouraging---PSO Complete performs slightly better.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere.eps}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations.  Our speculative algorithm has a swarm of 30 particles and does
  two iterations per round, while standard PSO algorithms use 240 processors
  and do one iteration per round.}
  \label{fig:basic-sphere}
\end{figure}

Next we look at the function Griewank, defined by the equation $f(\Vec{x}) =
\frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
\right) + 1$.  It is generally recommended to use the Ring topology when
optimizing the Griewank function, as Complete is prone to premature convergence
on a local optima.  We show results in \fig{basic-griewank1} for swarms of size
30 and 240 using the Ring topology. One can see in the figure that using
speculative evaluation initially outperforms PSO, but gets stuck in a local
optima and is eventually outperformed.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank1.eps}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations.  Our speculative algorithm has a swarm of 30 particles and does
  two iterations per round, while standard PSO algorithms use 240 processors
  and do one iteration per round.}
  \label{fig:basic-griewank1}
\end{figure}

\fig{basic-griewank1} is slightly deceptive, however, as with Griewank some of
the runs found the global optimum, 0, while some did not.  A more accurate plot
for this function is one showing what percentage of the runs have found the
global optimum at each iteration.  \fig{basic-griewank2} shows such a plot.  A
swarm of size 30 performing the PSO algorithm gets stuck a little less than
half of the time, while 240 particles provides enough exploration to find the
global optimum every time.  As would be expected, when our speculative
algorithm finds the global optimum, it finds it about twice as fast as regular
PSO---it is just performing regular PSO twice as fast with a smaller swarm
size.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank2.eps}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations.  Instead of showing average function value, we show the
  percentage of runs that have found the global optimum by each iteration.}
  \label{fig:basic-griewank2}
\end{figure}

Lastly we look at a function that does not show any kind of improvement with
this method.  The function Rastrigin is defined as $f(\Vec{x}) = \sum_{i=1}^D
\left(x_i^2 - 10\cos\left(2\pi x_i\right) + 10\right)$.  It has been shown that
with Rastrigin, the more particles there are in the swarm, the lower function
value it finds, up to at least 4000 particles~\cite{mcnabb-cec09}.  Smaller
swarms get caught in local optima.  Because our speculative algorithm has a
significantly smaller swarm size, it gets stuck at higher values while the
larger swarms performing regular PSO continue to improve the best value found.
\fig{rastrigin} shows the results graphically.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{rastrigin.eps}
  \caption{Function Rastrigin with a swarm that uses 240 processors per round
  of evaluations.  Our speculative algorithm has a swarm of 30 particles and
  does two iterations per round, while standard PSO algorithms use 240
  processors and do one iteration per round.}
  \label{fig:rastrigin}
\end{figure}

\section{Relaxing the Requirements}
\label{sec:relax}

While the idea of speculative evaluation in particle swarm optimization is
intriguing, the results of the previous section show that it requires too many
extra evaluations to produce competitive results.  On most functions, the
larger swarm size that can be used in regular PSO leads to better performance
than is found by trying to do speculative evaluation.  However, if we keep the
idea of speculative evaluation while relaxing the requirement of exactly
reproducing the behavior of the original PSO algorithm, we see some very
impressive results.

Should I present results for each improvement individually, or all together, or
both?

\subsection{Pick the Best Child}

In performing speculative evaluation as we have described it, $2(n+1)$
speculative evaluations are done per particle, while all but one of them are
completely ignored.  We can try to make use of those evaluations instead of
throwing them away.

@@

\subsection{Pruning the Speculative Evaluations}

Because the problem facing speculative evaluation is having too many possible
evaluations to perform, a natural step to take is to eliminate some of the
evaluations.  If we could reliably predict which branch were going to be taken,
we could limit ourselves to one speculative evaluation per particle instead of
$2(n+1)$.  While we cannot with certainty predict the correct branch every
time, we can look at the statistics of the behavior of particles in general and
perhaps come up with a good guess, reducing the number of necessary speculative
evaluations.

\subsubsection{Branch Statistics}

In Table~\ref{tab:evals} we presented all possible branches that a particle
with two neighbors could take.  Here we lump all of the neighbors together and
consider the statistics for the five branches shown in Table~\ref{tab:branches}.

\begin{table}
  \caption{Five Branches to Consider for Statistics}
  \label{tab:branches}
  \centering
  \begin{tabular}{rc|c}
	&$\pbest$ updated&Source of $\nbest$ update\\
	\hline
	1.&No update&No update\\
	2.&Updated&No update\\
	3.&Updated&The particle itself\\
	4.&No update&Some neighbor\\
	5.&Updated&Some neigbhor\\
  \end{tabular}
\end{table}

\begin{table}
  \caption{Branch Statistics in PSO}
  \label{tab:stats}
  \centering
  \begin{tabular}{c|c|c|c|c|c|c}
	Topology&Function&1&2&3&4&5\\
	\hline
	Ring&Sphere&0.530&0.093&0.114&0.202&0.062\\
	&Griewank&0.517&0.084&0.122&0.207&0.070\\
	&Rastrigin&0.495&0.048&0.146&0.213&0.099\\
	&Rosenbrock&0.513&0.074&0.129&0.211&0.073\\
	\hline
	Random&Sphere&0.667&0.119&0.026&0.156&0.031\\
	&Griewank&0.690&0.109&0.025&0.149&0.027\\
	&Rastrigin&0.819&0.055&0.015&0.100&0.010\\
	&Rosenbrock&0.742&0.077&0.022&0.140&0.018\\
	\hline
	Complete&Sphere&0.319&0.092&0.002&0.451&0.135\\
	&Griewank&0.353&0.084&0.002&0.441&0.119\\
	&Rastrigin&0.477&0.067&0.002&0.382&0.070\\
	&Rosenbrock&0.353&0.034&0.003&0.544&0.066\\
  \end{tabular}
\end{table}

Row 1 in Table~\ref{tab:branches} corresponds to the phenomon commonly called
stagnation in the literature, but only with a single particle.  As an
interesting aside, we found that complete stagnation practically never occurs
in PSO, though frequently large percentages of the swarm are stagnant.
Iterations where the entire swarm is stagnant average less than 2 in 1000
iterations, unless the swarm converges beyond machine precision.  When graphs
of the best function value flatline, what really has happened is a contraction
of the particles' velocity, not stagnation in the technical sense.

We now show with what percentage a particle takes each of these branches for
three different topologies and four different functions.  All of our statistics
are from swarms of 240 particles.  Brief experimentation showed that other
swarm sizes had similar statistics.  We ran 750 iterations on all combinations
of functions and topologies except for the functions Griewank and Rastrigin
with the Complete topology.  We found that those runs frequently converged past
machine precision after 500 iterations, and that led to erroneously high values
for the probability of stagnation.  Instead we ran for only 450 iterations on
those two combinations.  All of our results were averaged over 20 runs of the
algorithm; thus the probabilities presented are the averages of 3.6 million
trials for the branch taken (2.16 million for the two with only 450
iterations).  Table~\ref{tab:stats} contains the results, with the branch
number coming from Table~\ref{tab:branches}.  We have already defined the
functions Sphere, Griewank, and Rastrigin.  Rosenbrock is defined as
$f(\Vec{x}) = \sum_{i=1}^D \left(100(x_{i+1}-x_i^2)+(x_i-1)^2\right)$.

The probabilities presented in Table~\ref{tab:stats} are interesting in and of
themselves and could probably be used to better understand the characteristics
of various topologies.  It is notable that there is small variation between
functions in any given topology, but the variation across topologies is far
greater.  However, our concern is with speculative evaluation.  We are
interested in predicting the branch that any given particle will take at a
particular iteration.

From Table~\ref{tab:stats} we can see that with the Random topology, we can
pick the first branch, corresponding to stagnation, and be right around 70\% of
the time.  With the Ring topology, we would be right 50\% of the time.
Branches four and five really correspond to several actual branches, as all of
the neighbors are lumped together.  The 20\% probability of taking branch four
with the Ring topology can be split into two branches, as there are only two
neighbors.  It also turns out that the neighbor that last updated the
neighborhood best is the most likely to update it next time, so keeping track
of that information could be fruitful in trying to predict that branch.

The statistics for the Complete topology are less promising (as there are 240
neighbors that branch four splits into, instead of two), though we do not use
that topology with speculative evaluation anyway.

Because there are a few branches with very high probabilities, we can have hope
that cutting out some of the evaluations that have low probability will lead to
an increase in overall performance.

\subsubsection{What if we guessed wrong?}

When not all of the branches are evaluated, there is some probability of not
evaluating the branch that was actually taken by the original particle.  We 
need to decide what to do in this case.

One possibility is to leave the particle as it is, not accepting any of the
speculative evaluations, because none of the were correct.  This leads to
particles being at different iterations, as some particles guess correctly
while others do not.  Thus we lose exact compatibility with the original PSO,
though this particular relaxation is nothing new; asynchronous adaptations of
PSO do the same thing.  As an aside, it is equivalent in this case to simply
increment the iteration number of particles which fail to correctly predict
their branch.  This keeps the iteration number constant across all particles,
simplifying the work needed to be done when dynamic topologies are involved.

Another possibility is to pick the best child, as described above.  This
ignores the fact that the branch might have been wrong; we do not care, because
we just want the child that had the best value, anyway.  In almost all of the
experiments that we ran, it turned out that picking the best child performed
better than keeping the particle back an iteration.

\subsection{More than one iteration ahead}

\subsection{Tokens}

@@

@@

@@

This work represents the first attempt of which we are aware at using
additional processors in parallel with PSO to do something other than either
add particles to the swarm or parallelize the evaluation of the objective
function.  We anticipate that many future advancements can be made.

We have speculated (pardon the pun\dots) that our ``pick the best child''
method simply draws a number of samples from the next iteration of the sampling
distribution of PSO and keeps one of them.  It would then be expected that the
sampling distribution of that method is very similar to the distribution of the
original PSO.  This remains to be proven, or even demonstrated empirically.

We gave a passing glance to some very interesting statistical characteristics
of PSO.  Our motivation was to use the statistics to improve our speculative
algorithm, but it seems there are some nuggets in the numbers that could be
used to improve our understanding of the behavior of PSO in general, and what
role a topology plays.  A potentially very fruitful avenue of future work would
be to further analyze the branch statistics in PSO.  There are some particles
that seem to drive the swarm and others that are stagnant almost the entire
time.  Such information could be used in a TRIBES-like fashion to determine
which particles should be removed from the swarm, perhaps starting them close
to well-performing particles.

We turned what used to be a very simple algorithm with only one major parameter
to tweak (the topology) into a big hairy mess with countless possibilities for
speculative evaluations.  Perhaps future research will be able to narrow down
the endless possibilities into the few best ones, making the algorithm simple
again.
